Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-11 write-ups/PLDI/paper.tex
Spelling

An OpenCL program is divided into \emph{host} and \emph{device}
components.  A host application runs on the CPU and launches one or
more \emph{kernels} that run on accelerator devices---GPUs in the
context of this paper.  A kernel is writen in OpenCL C, based on C99.
All threads executing a kernel execute the same entry function with
identical arguments, but a thread may invoke the $\getglobalid()$
function to query its unique thread id in order to access different
data or follow different control flow paths to other threads.

@@

The $\offerkill$ primitive allows the cooperative kernel to return
compute units to the scheduler by offering to sacrifice workgroups.
The idea is that while having the scheduler be permitted to terminate
the execution of workgroups in an arbitary fashion would be drastic,
the kernel may contain identifiable points of execution at which a
workgroup could gracefully leave the computation.  We give examples of
these for work stealing and graph traversal below.

@@

\subsection{Programming With Coopeative Kernels}\label{sec:programmingguidelines}

@@

For the work stealing example, there is no state to transmit since a
computation is entirely parametrized by a task, which is retrieved from
a task pool located in global memory. Care must be taken on the dynamic
change of the number of workgroups though: we cannot assume that the
total number of task pools is equal to $\getnumgroups$. Thus, the number
of pools must be given as an argument, and the $\mathsf{pool}$ argument
must be obtained from the group id modulo the total number of
pools. This total number of pools must also be passed to
$\mathsf{pop\_or\_steal}$, since when a workgroup wants to steal it must
access other pool indexes, which are bounded by the total number of
pools.

@@

In summary, our experience so far indicates that cooperative kernels
is a good fit for current irregular algorithms.  Of cousre this is, in
part, because we designed the model with these algorithms in mind, but
we argue that the fit is natural not just for the examples that we
study, but more generally for at least these two patterns of irregular
algorithm.

@@

Getting this non-functional requirement right depends on the
constraints of the tasks to be co-scheduled with a cooperative kernel,
on properties of the cooperative kernel itself, and on the efficiency
of the GPU that is used for execution.  In \mysec\ref{sec:representativeworkloads} we conduct
experiments to understand the response rate that would be required to
co-schedule graphics rendering with a cooperative kernel in order to
achieve a smooth frame rate, and we evalaute responsiveness in
practice in \mysec\ref{sec:responsiveness}.

@@

Recall that the cooperative kernel would ideally utilise the $N$
workgroups requested upon kernel launch, in order to conduct its
processing efficiently.  The scheduler should thus aim to provide with
cooperative kernel with as many workgroups, up to this number, as
other constraints allow.  As a result, the scheduler should generally
accept an $\offerkill$ only if a comput unit is required for another
task, and should respond positively to $\offerfork$ calls by
scheduling new workgroups if compute units are available.  Our
implementation of cooperative scheduling (\mysec\ref{sec:implementation}) is as
generous as possible to the cooperative kernel in this regard.
However, these are not hard-and-fast rules: a scheduler might employ
heuristics based on historical workloads to avoid thrashing.  For
example, if graphics calls are known to arrive at a certain rate then
the scheduler might decide that it is worth accepting an $\offerkill$
from the cooperative kernel on the assumption that a graphics task
will be requested momentarily.  Similarly, the scheduler might provide
some, but not all, compute units to the cooperative kernel in response
to an $\offerfork$ call if it has reason to believe that the remaining
compute units are likely to be useful for other high priority tasks
that are likely to arrive imminently.

@@

\subsection{The Overhad of Moving to Cooperative Kernels}

@@

\cite{DBLP:conf/ics/WuCLSV15} \ADComment{Basic idea: map tasks to
  SMs, not to workgroups, so that tasks taht would benefit from being
  processed by the same SM can be scheduled to the same SM.  Has
  little to do with preemption, but more to do with irregular
  computation.}

>>>>

An OpenCL program is divided into \emph{host} and \emph{device}
components.  A host application runs on the CPU and launches one or
more \emph{kernels} that run on accelerator devices---GPUs in the
context of this paper.  A kernel is written in OpenCL C, based on C99.
All threads executing a kernel execute the same entry function with
identical arguments, but a thread may invoke the $\getglobalid()$
function to query its unique thread id in order to access different
data or follow different control flow paths to other threads.

@@

The $\offerkill$ primitive allows the cooperative kernel to return
compute units to the scheduler by offering to sacrifice workgroups.
The idea is that while having the scheduler be permitted to terminate
the execution of workgroups in an arbitrary fashion would be drastic,
the kernel may contain identifiable points of execution at which a
workgroup could gracefully leave the computation.  We give examples of
these for work stealing and graph traversal below.

@@

\subsection{Programming With Cooperative Kernels}\label{sec:programmingguidelines}

@@

For the work stealing example, there is no state to transmit since a
computation is entirely parameterised by a task, which is retrieved from
a task pool located in global memory. Care must be taken on the dynamic
change of the number of workgroups though: we cannot assume that the
total number of task pools is equal to $\getnumgroups$. Thus, the number
of pools must be given as an argument, and the $\mathsf{pool}$ argument
must be obtained from the group id modulo the total number of
pools. This total number of pools must also be passed to
$\mathsf{pop\_or\_steal}$, since when a workgroup wants to steal it must
access other pool indexes, which are bounded by the total number of
pools.

@@

In summary, our experience so far indicates that cooperative kernels
is a good fit for current irregular algorithms.  Of course this is, in
part, because we designed the model with these algorithms in mind, but
we argue that the fit is natural not just for the examples that we
study, but more generally for at least these two patterns of irregular
algorithm.

@@

Getting this non-functional requirement right depends on the
constraints of the tasks to be co-scheduled with a cooperative kernel,
on properties of the cooperative kernel itself, and on the efficiency
of the GPU that is used for execution.  In \mysec\ref{sec:representativeworkloads} we conduct
experiments to understand the response rate that would be required to
co-schedule graphics rendering with a cooperative kernel in order to
achieve a smooth frame rate, and we evaluate responsiveness in
practice in \mysec\ref{sec:responsiveness}.

@@

Recall that the cooperative kernel would ideally utilise the $N$
workgroups requested upon kernel launch, in order to conduct its
processing efficiently.  The scheduler should thus aim to provide with
cooperative kernel with as many workgroups, up to this number, as
other constraints allow.  As a result, the scheduler should generally
accept an $\offerkill$ only if a compute unit is required for another
task, and should respond positively to $\offerfork$ calls by
scheduling new workgroups if compute units are available.  Our
implementation of cooperative scheduling (\mysec\ref{sec:implementation}) is as
generous as possible to the cooperative kernel in this regard.
However, these are not hard-and-fast rules: a scheduler might employ
heuristics based on historical workloads to avoid thrashing.  For
example, if graphics calls are known to arrive at a certain rate then
the scheduler might decide that it is worth accepting an $\offerkill$
from the cooperative kernel on the assumption that a graphics task
will be requested momentarily.  Similarly, the scheduler might provide
some, but not all, compute units to the cooperative kernel in response
to an $\offerfork$ call if it has reason to believe that the remaining
compute units are likely to be useful for other high priority tasks
that are likely to arrive imminently.

@@

\subsection{The Overhead of Moving to Cooperative Kernels}

@@

\cite{DBLP:conf/ics/WuCLSV15} \ADComment{Basic idea: map tasks to
  SMs, not to workgroups, so that tasks that would benefit from being
  processed by the same SM can be scheduled to the same SM.  Has
  little to do with preemption, but more to do with irregular
  computation.}
